Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-05-15 trunk/papers/ArticleZ80/source/paper.tex
novos pequenos ajustes

\title{Experience in Modeling a Microcontroller Instruction Set Using B}
%
\hyphenation{se-pa-ra-te}
\hyphenation{in-struc-tion}
\hyphenation{re-gis-ter}
\hyphenation{re-gis-ters}
%
%
%
\author{Val\'{e}rio Medeiros Jr\inst{1}, David D\'{e}harbe\inst{2}}

\institute{ Federal Institute of Education, Science and Technology of Rio Grande do Norte, Natal RN 59015-000, Brazil
           \and 
		    Federal University of Rio Grande do Norte, Natal RN 59078-970, Brazil}

@@

The rigorous development of an embedded software in a microcontroller requires a
specification of its instruction set and the software requirements specification.
The actual industrial practice is to document the instruction set informally
using natural language or with added semi-formal notation. This practice leaves
space for errors and inconsistencies \cite{springerlink:Yuan2011}. This paper
presents specifically a B formal model of the Z80 instruction set. This model 
can be  used to document, simulate, test and in a model-driven approach to
design and develop software from the functional specification level down to
assembly.
% Using this model-driven approach and the instruction set model, a small case
% study to analyse a petroleum production test system in oil field was developed.
%  \paragraph[Keywords]{ \textbf{Keywords}: Embedded Software, Simulation, B
% method and Verification%}
\end{abstract}

%\keywords{ Validation, Verification, Microcontroller, Hardware}
%
%\setcounter{secnumdepth}{1}
%\setcounter{chapter}{1}
%\setcounter{section}{1}

@@

The new model supports animation in ProB\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to
  visit our repository at: http://code.google.com/p/b2asm/.}  no
longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of microcontroller model. Such
library has many definitions of concepts common in microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using Z80 assembly language to analyse the code simulation
and the verification process.

Providing the formal model of an instruction set yields many
benefits. The animation of Z80 model allows to simulate the execution
assembly programs including support for interrupts and input and
output instructions. Other possible uses of a formal model of a
microcontroller instruction set include documentation, the
construction of simulators, and can be possibly the starting point of
a verification effort for the actual implementation of a Z80
design. Moreover, the model of the instruction set could be
instrumented with non-functional aspects, such as the number of cycles
it takes to execute an instruction, to prove lower and upper bounds on
the execution time of a routine.  Two aspects of these formalization
are of particular importance to us: provide a more solid documentation
artifact for microcontrollers and build a reference model for a formal
compilation approach in B.

\subsection{Problems in formalizing instruction set}
% Compact version

Many times, each instruction of microcontrollers is shown in its official manual
using textual description, math description, examples, encoding, number of cycles
and other information. Sometimes, the instruction descriptions have
an own notation, some informations are organized in different sections
of document and the textual description are not standardized. 
The manuals also have semi-formal and informal descriptions, these descriptions
may leave errors, inconsistency and ambiguity.  
For example in official manual of the microcontroller Z80 \cite{Z80_manual}, if the users  
want to know the action of an instruction and its effects on the flag register then he needs
to read an informal textual description, that  sometimes can be placed in different pages. 
Furthermore, the official manual of the microcontroller Z80 has also several
problems that were identified over time~\cite{UndocumentedZ80}.

A solution to avoid ambiguity and inconsistencies is to specify
formally the assembly instruction set, this creates a pattern of
representation and validates properties of instructions. The formal
model also restricts the definitions to correct typing and uses only
well-defined expressions. Furthermore, the developer can use with B
method different levels of abstraction and add more specific details
in refined models.

@@

%[NEW PARAGRAPH-TO-REVIEW]
%Similarly, several other general functions and lemmas were created for all other data types.
%Some times, we need create arithmetic and logic functions that are more specific for a
%determined computer architecture. These functions are developed in the module ALU 
%(arithmetic  logic unit). The following shows some examples of specific functions used in
%microcontroller model of Z80. For example, the function $\textit{instruction_next}$
%receives a $\textit{BYTE}$ and returns a $\textit{BYTE}$ that contains a sequence of bit:
%a signal bit,

Nearly 50\% of all these proof obligations shown in the
Table~\ref{tab:types} are verified automatically or with few user
commands. However many others proof obligations are difficult to
verify, as they involve functions that contain arithmetic expressions
that are difficult to manipulate with the interactive prover of
AtelierB.

\subsection{Verification} 
\label{sec:VerificationHardwareLibrary}

This section presents the verification of the aforementioned modules
using AtelierB provers.  By default, AtelierB~\cite{atelierB} does
not generate proof obligations of the type of functions defined in the
 $PROPERTIES$ clause of a machine. For example, the definition $inc \in
\{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x + 1)$,
does not create a proof obligation to check the consistency of the
definition. However, this definition is inconsistent since $inc(1)=2$
and $ 2 \not\in \{0,1\}$.

To avoid such inconsistencies, we state separately the required properties in the
\textit{ASSERTIONS\/} clause, that cause corresponding proof obligations to be
generated and verified. However, we have not been able to build the corresponding
proofs using the base of rules included in the free distribution of AtelierB. So
we formalized new proof rules to check these properties in AtelierB and used them
in the interactive prover of AtelierB.

%\marginpar{Não seria interessante ter colocado
%  esta figura mais cedo? Na seção 2.5, por exemplo? E no texto da
%  seção 2.6 poderia enumerar os poucos casos verificados sem regras
%  adicionais.}
The hardware library has many conversion functions with
arithmetic,  for example, Bit\_vector to Natural, Natural to
Bit\_vector, Integer to Bit\_vector and Bit\_vector to Integer. These
functions are organized in modules and each module contains auxiliary
functions for a data type. The modules are represented by gray
rectangles in Figure \ref{DiagramTypesRules}.

%and they have the mean: \textit{BYTE} is a bit vector with eight elements enumerated
%starting from one; \textit{UCHAR} is a range between 0 and 255; \textit{SCHAR} is a range between -128 and 127; 
%\textit{BV16} is a bit vector with sixteen elements enumerated starting from one; \textit{USHORT} is a range 
%between 0 and 65535; and \textit{SSHORT} is a range between -32768 and 32767.

@@

%\textbf{We needed to verify the type of these conversion
%  functions}.\marginpar{Não precisa repetir
%  isto.} % There are twenty rules associated this properties
The conversion functions need to satisfy two properties: to be
bijective and to be equal to the reverse of the dual conversion
function. The three simplest proof obligations were verified
automatically with AtelierB: (\it uchar\_schar\rm \ $\in$
\textit{UCHAR} $\bij$ \textit{SCHAR}), (\it schar\_uchar\rm \ $\in$
\textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it byte\_bv16\rm \ $=$ \it
bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we again had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

The Z80 has 158 different instructions, including all the 78 from Intel 8080
microprocessor, and all of them were formally specified in B. These instructions are classified
into these categories: load and exchange; block transfer and search; arithmetic
and logical; rotate and shift; bit manipulation; jump, call and return;
input/output; and basic cpu control. Each category of instruction has different
elements of specification.

This section shows the elements that make up the different types of instructions.
The main elements are specified in the microcontroller module Z80 and parts of it
are presented below.  Basically, the state of microcontroller is formed by data 
from program counter, ports, registers, etc. The transitions between the states 
are actioned by execution of instruction or an external action. Groups of registers
are represented by the variables of the specification states that appear in clause
\textit{VARIABLES}. The declaration of valid states of variables is
in the \textit{INVARIANT} clause and the initial state is defined in the
\textit{INITIALISATION} clause. The assembly instructions are defined by the
clause \textit{OPERATIONS} and, in general, each instruction has three actions: update 
the program counter, update the flag register and its main effect.
%Several functions were created to Z80.
General functions that can be used in others microcontroller models are defined in
the modules of data types. Specific functions of the microcontroller are defined in the ALU (arithmetic logic unit) module.

@@

%\subsection{Considerations about the Z80 model}

The Z80 formal model provides many benefits, because of the verification of
generated proof obligations guarantees: the correct use of data types, the developed
security properties  and the well-defined of all the expressions. %\footnote{An expression is called ``well-defined'' (or unambiguous) if its definition assigns it a unique interpretation or value.}.
Furthermore, the designer has a big flexibility to create new and specific security properties, this is very useful
to adjust the verification in accordance with the requirements. Moreover, this
example of model could replace or improve the used documentation for users and
assembly programmers. Z80 model is also useful to develop verified software 
up to the assembly level. %There are many benefits when a formal model is
%developed, but to verify the model is not an easy task, the first version of Z80 model have already verified %completely,but the new Z80 model with support to animation in ProB has not been verified completely,
%because the changes in Z80 model are recent and the verification time is long.

@@

There are in the literature several similar approaches to model hardware and instruction set using B.
The paper \cite{springerlink:Yuan2011} reports a method to specify, design and construct sound and complete instruction set model by stepwise refinement and formal proof using Event-B. It also discusses about desirable properties of instruction set model. However, our work \cite{Valerio_SBMF09} and \cite{Subotic2010} use the B method, that seems quite appropriated to software development, because it has 
an implementable language defined, called B0, a model-driven approach \cite{Dantas_SBMF08} to develop software from the functional specification level down to assembly, and tools to convert the models to a programming language. An related experience on using B in the design of secure micro-controllers is present in \cite{Marc20113}. It tries to show the feasibility of such a technique for high confidence trustful devices. The work \cite{Subotic2010} describes a similar approach with MIPS CPU architecture and it shows how the CPU is formally specified and implemented using the B method.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the instruction set of
microcontroller using the B method.  During the construction of this model, some
problems were found in the official reference for Z80
microcontroller~\cite{Z80_manual}. We have consulted different sources \cite{Simulator_z80,UndocumentedZ80,Z80_manual}  to avoid problems in modeling and have developed a case study: a verified embedded software.
This case study was interesting to analyse the technique of formal verification up to assembly level.

% TODO: Adaptar para apenas fundamentar melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The following works quoted are directly related to the objective this paper.
The first work \cite{DantasSemish2008} shows the developed metodology to verify
software up to assembly level using B. A second work\cite{Dantas_SBMF08}
presents more details about the verification approach and a small 
software that was verified up to assembly level in three different platforms. 
The last work \cite{Valerio_SBMF09} shows a general view of the first version of Z80
model and techniques used in verification process. Basically, the new Z80 model presented here has added support for animation in ProB and the representation of external actions (interruptions).
%Link com o parágrafo anterior 
These works created an important step to software verification up to assembly
language, they show the current difficulties and suggest improvements in the
B tools.

% TODO
% [Trabalhos futuros ] LLVM:Existe também a pretensão de especificar as

Future works comprise the development of software with the B method from
functional specification to assembly level, using the Z80 model presented in this
work. The mechanic compilation from B algorithmic constructs to assembly platform
is also envisioned. Finally, another important activity is to develop a formal
model of a platform used in LLVM Compiler~\cite{DBLP:conf/cgo/LattnerA04}.

>>>>

\title{Experience in Modeling a Microcontroller Instruction Set Using B}
%
\hyphenation{se-pa-ra-te}
\hyphenation{in-struc-tion}
\hyphenation{re-gis-ter}
\hyphenation{re-gis-ters}

%
%
%
\author{Val\'{e}rio Medeiros Jr\inst{1}, David D\'{e}harbe\inst{2}}

\institute{Federal Institute of Education, Science and Technology of Rio Grande do Norte, Natal, RN, Brazil
  \and 
  Federal University of Rio Grande do Norte, Natal, RN, Brazil}

@@

  This paper presents a formal model, developed in B, of the Z80
  instruction set. The formal model includes different libraries that
  specify entities commonly used in microcontrolers and
  microprocessors. Not only this model can be used for documentation
  purposes, but also to simulate and test the assembly code. It is also
  a fundamental component to extend the reach of the B method from the
  specification of the functional requirements of a system to an assembly-level
  implementation thereof.
% Using this model-driven approach and the instruction set model, a small case
% study to analyse a petroleum production test system in oil field was developed.
%  \paragraph[Keywords]{ \textbf{Keywords}: Embedded Software, Simulation, B
% method and Verification%}
\end{abstract}

@@

The new model supports animation in ProB~\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to
  visit our repository at: http://code.google.com/p/b2asm/.}  no
longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of the microcontroller model. Such
a library has many definitions of concepts common to microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using the Z80 assembly language to analyse the code simulation
and the verification process.

The formal model of an instruction set has several
applications. Animation allows to simulate the execution of assembly
programs including support for interrupts, input and output
instructions. Other possible uses include documentation, the
construction of simulators, and can possibly be the starting point of
a verification effort for the actual implementation of a Z80
design. Moreover, the model of the instruction set could be
instrumented with non-functional aspects, such as the number of cycles
it takes to execute an instruction, to prove lower and upper bounds on
the execution time of a routine.  Two aspects of such formalizations
are of particular importance to us: provide a more solid documentation
artifact for microcontrollers and build a reference model for a formal
compilation approach in B.

It is often the case that the official documentation of an instruction
uses textual description, mathematical formulas, examples,
etc. Sometimes, the instruction descriptions have a specific notation,
some informations are organized in different sections of a document and
the textual descriptions are not standardized.  The manuals also have
semi-formal and informal descriptions, these descriptions may leave
errors, inconsistency and ambiguity.  For example, in the official
manual of the microcontroller Z80 \cite{Z80_manual}, the semantics of
an instruction (in particular its effects on the flag register) are
described informally, and this description is scattered in
different pages.  Furthermore, the official manual of the
microcontroller Z80 has also several problems that were identified
over time~\cite{UndocumentedZ80}.

A solution to avoid ambiguity and inconsistencies is to formally
specify the assembly instruction set, making it possible to prove
properties of the instructions. The formal model also restricts the
definitions to correct typing and uses only well-defined
expressions. Furthermore, the developer may use the refinement
capabilities of the B method to specify at different levels of
abstraction and add more specific details or useful restrictions in
refined models.

@@

%TODO:[NEW PARAGRAPH-TO-REVIEW]
Similarly, several other general functions and lemmas were created for all other data types.
Some times, we need  create arithmetic and logic functions that are more specific for a
determined microprocessor, like Z80 that is presented in the section \ref{sec:z80}.

%TODO:[NEW PARAGRAPH-TO-REVIEW]
The definitions, functions and properties from hardware library and types library are difficult to verify,
on the other hand  these libraries can be reused in several others  microprocessors without to verify again.
Nearly 50\% of all these proof obligations shown in the Table~\ref{tab:types} are verified automatically 
or with few user pass\footnote{The user pass are steps that aid the prover to find the proof.} .
But many others proof obligations are difficult to verify, because they involve functions
that contain arithmetic expressions that are complex to manipulate with the interactive
prover of AtelierB.  However, all theses 531 proof obligations were verified completely.

\subsection{Verification of hardware library and types library} 
\label{sec:VerificationHardwareLibrary}

This section presents the verification of the aforementioned modules
using AtelierB provers.  By default, AtelierB~\cite{atelierB} does not
generate proof obligations to check the type of functions defined in
the $PROPERTIES$ clause of a machine. For example, the definition $inc
\in \{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x +
1)$, does not create a proof obligation to check the consistency of
the definition. However, this definition is inconsistent since
$inc(1)=2$ and $ 2 \not\in \{0,1\}$.

To avoid such inconsistencies, we state separately the required
properties in the \textit{ASSERTIONS\/} clause: this causes
corresponding proof obligations to be generated and verified. However,
we have not been able to build the corresponding proofs using the 
rule base included in the free distribution of AtelierB. So we
formalized new proof rules to check these properties in AtelierB and
used them in the interactive prover of AtelierB.

The hardware library has many conversion functions with
arithmetic,  for example, Bit\_vector to Natural, Natural to
Bit\_vector, Integer to Bit\_vector and Bit\_vector to Integer. These
functions are organized in modules and each module contains auxiliary
functions for a data type. The modules are represented by gray
rectangles in Figure \ref{DiagramTypesRules}.

@@

The conversion functions need to satisfy two properties: bijectivity
and equality to the reverse of the dual conversion function. Three
such proof obligations were verified automatically with AtelierB: (\it
uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR}), (\it
schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it
byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we again had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

The Z80 has 158 different instructions, including all the 78 from the Intel 8080
microprocessor, and all of them were formally specified in B. These instructions are classified
into these categories: load and exchange; block transfer and search; arithmetic
and logical; rotate and shift; bit manipulation; jump, call and return;
input/output; and basic cpu control. Each category of instruction has different
elements of specification.

This section shows the elements that make up the different types of
instructions. The main elements are specified in the microcontroller module Z80
and parts of it are presented below.  Basically, the state of microcontroller is
formed by data from program counter, ports, registers, etc. The transitions
between the states are actioned by execution of instruction or an external
action. Groups of registers are represented by the variables of the
specification states that appear in clause \textit{VARIABLES}. The declaration
of valid states of variables is in the \textit{INVARIANT} clause and the initial
state is defined in the \textit{INITIALISATION} clause. The assembly
instructions are defined by the clause \textit{OPERATIONS} and, in general, each
instruction has three actions: update the program counter, update the flag
register and its main effect.
% Several functions were created to Z80.
General functions that can be used in other microcontroller models are defined
in the modules of data types. Specific functions of the microcontroller are
defined in the ALU (arithmetic logic unit) module.

@@

\subsection{Verification process}% ESTOU TRABALHANDO AQUI
% OK -> como as regras foram utilizadas ?  Respondido na seção das regras
% TODO:ENGLISH REVIEW parágrafo inicial: VISÃO GERAL
This project has many proof obligations and its verification can be a hard task,
hence this section explains about the verification process of proof
obligations and its importance. % [O que as provas garantem? citar que énecessário buscar técnicas eficientes para realizar as provas ]
The proof obligations allow to verify the data types, important system properties and if
the expressions are well-defined (WD)\footnote{An expression is called ``well-defined''
(or unambiguous) if its definition assigns it a unique interpretation or value.}. The
properties provide additional guarantees because they can set many safety
rules. However, the model can be very difficult to prove.

 % -> O que significa cada coluna e linha da tabela ?
The Table~\ref{tab:completestatistcs} contains some statistics of verification
process, it shows only the number of no obvious proof
obligations\footnote{Obvious proof obligations are simple formulas automatically eliminated by
the tool.} and the titles of columns use acronyms: {\small \textbf{PO}} - number
of proof obligations; {\small \textbf{UPO}} - number of unproved proof
obligations; {\small \textbf{WD PO}} - number of proof obligations  related to
well definition lemmas; {\small \textbf{UWD PO}} - number of unproved proof
obligations related to well definition lemmas.

\begin{table} [h]
\caption{Verification statistics }
\label{tab:completestatistcs}
\begin{center} 
\begin{tabular}{|p{3.1cm}|c|c||c|c||c|c|c|}
\hline 
{\small \textbf{Component}}&{\small \textbf{PO}}&{\small \textbf{UPO}}&{\small \textbf{WD PO}}&{\small \textbf{UWD PO }}&{\small \textbf{Total PO}}&{\small \textbf{Total UPO}}& {\small \textbf{Total Percentage}}  \\\hline
BIT&49&0&69&0&118&0&100\%\\\hline
BV16&6&0&69&0&75&0&100\%\\\hline
BYTE&18&0&136&0&154&0&100\%\\\hline
POWER&3&0&4&0&7&0&100\%\\\hline
POWER2&18&0 &0&0&18&0&100\%\\\hline
SCHAR&4&0&26&0&30&0&100\%\\\hline
SSHORT&4&0&58&0&62&0&100\%\\\hline
TYPES&71&0&0&0&71&0&100\%\\\hline
UCHAR&4&0&26&0&30&0&100\%\\\hline
USHORT&4&0&58&0&62&0&100\%\\\hline
ALU&51&22&117&47&168&69&59\%\\\hline
MEMORY&13&0&0&0&13&0&100\%\\\hline
Z80 -  {\scriptsize Initialization, Properties, Assertions} &66&9&150&11&216&20&91\%\\\hline
Z80 - {\scriptsize Input and Output}&46&11&281&16&327&27&92\%\\\hline
Z80 - {\scriptsize Logic Arithmetic 1}&88&33&343&84&431&117&73\%\\\hline
Z80 - {\scriptsize Logic Arithmetic 2}&40&20&178&27&218&47&79\%\\\hline
Z80 - {\scriptsize Bit Control}&111&34&763&157&874&191&78\%\\\hline
Z80 - {\scriptsize External Actions}&31&0&60&38&91&38&59\%\\\hline
Z80 - {\scriptsize General 1}&81&19&420&20&501&39&92\%\\\hline
Z80 - {\scriptsize General 2}&26&10&139&14&165&24&86\%\\\hline

\textbf{Total}&\textbf{734}&\textbf{158}&\textbf{2897}&\textbf{364}&\textbf{3631}&\textbf{569}&\textbf{84}\%\\\hline

\end{tabular}
\end{center} 
\end{table}

%->Explicaçao de mais detalhes da tabela.

% Porque o modelo não foi verificado completamente ?
There are many benefits when a formal model is developed, but to verify the
model is not an easy task, the first version of Z80 model have already verified
completely. However, this new Z80 model with support to animation in ProB has not
been verified completely because the changes in Z80 model are recent and the
verification time is long.
% ->Quanto tempo foi gasto nessa verificação ?  Aproximadamente quanto tempo de
% processamento elas reduziram ?
Nevertheless this project has 3631 proof obligations and the first version of
Z80 model spent approximately two months of work to finish the verification
process. The same  developed verification strategies were adjusted and reused in
new Z80 model, so approximately  7 days of work was enough to verify 3062  proof
obligations (84\%) and remaining only 569 proof obligations (16\%) not yet
verified.

Two development strategies were important to improve the verification of 3062
proof obligations: the intensive use of parallelism and user pass. The Z80 module
was decomposed in 8 modules of instructions, it was very helpful to decrease the
time for verification. Because AtelierB can run simultaneously \textit{N} instances of 
provers, one per module, reducing the time verification by up to \textit{N}  times when
a processor with \textit{N} cores is used  or the process is submitted to \textit{N} 
networked computers. So we also set up a proving environment consisting of networked
computers to take advantage of the distribution facilities provided in the B development
environment.
% Resumo de como aconteceu o processo de verificação,
A second interesting strategy  was to reuse the user pass. Usually, an used
user pass to aid the verification of one proof obligation can solve several
similar proof obligations. Thus, few generic user pass are needed to prove many
proof obligations. As there are many similar assembly instructions, some
human-directed proofs, when replayed, could discharge other proof obligations. A
good example is a set of 17 user pass that quickly aided the verification of
approximately 2533 of WD proofs.

\subsection{Considerations about the Z80 model}

The Z80 formal model provides many benefits, because of the verification of
generated proof obligations guarantees: the correct use of data types, the developed
security properties  and the well-defined of all the expressions. 
Furthermore, the designer has a big flexibility to create new and specific security properties, this is very useful %TODO: COLOCAR APÓS TODOS OS BENEFICIOS PROPORCIONADOS COM A SIMULAÇÃO e COMPATIBILIDADE COM O PROB. (OBTER DO PARAGRAFO ANTERIOR DA SEÇAO VERIFICATION PROCESS).
to adjust the verification in accordance with the requirements. Moreover, this
example of model could replace or improve the used documentation for users and
assembly programmers. Z80 model is also useful to develop verified software 
up to the assembly level.

@@

There exists several approaches to model hardware and instruction set
using B a.  The paper \cite{springerlink:Yuan2011} reports a method to
specify, design and construct sound and complete instruction set model
by stepwise refinement and formal proof using Event-B. It also
discusses desirable properties of the instruction set model. However,
our work \cite{Valerio_SBMF09} and \cite{Subotic2010} use the B
method, that seems quite appropriated to software development, because
it has an implementable language defined, called B0, a model-driven
approach \cite{Dantas_SBMF08} to develop software from the functional
specification level down to assembly, and tools to convert the models
to a programming language.

A related experience on using B in the design of secure
micro-controllers is present in \cite{Marc20113}. It tries to show the
feasibility of such a technique for high confidence trustful
devices. The work \cite{Subotic2010} describes a similar approach with
MIPS CPU architecture and it shows how the CPU is formally specified
and implemented using the B method.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the
instruction set of microcontroller using the B method.  During the
construction of this model, some problems were found in the official
reference for Z80 microcontroller~\cite{Z80_manual}. We have consulted
different sources \cite{Simulator_z80,UndocumentedZ80,Z80_manual} to
avoid such problems in modeling and have developed a case study: a
verified embedded software.  This case study was interesting to as a
feasibility analysis of the approach to derive formally assembly code
from a specification of functional requirements in the B method.

%TODO: COLOCAR AQUI OS PRINCIPAIS BENEFÍCIOS PROPORCIONADOS PELO TRABALHO

% fundamenta melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The following works are directly related to the objective this paper.
The approach to develop verified software down to the assembly level
using B was described in~\cite{DantasSemish2008}. A first case study
for this approach was reported in~\cite{Dantas_SBMF08}, presenting
more details as well as a small software that developed up to assembly
level in three different platforms. A general view of a previous used
in verification process were presented in~\cite{Valerio_SBMF09}. The
model presented in the current paper added support for animation in
ProB and the representation of external actions (interruptions).
%Link com o parágrafo anterior 
This set of papers present an important step to software verification
up to assembly language, showing the current difficulties and
suggesting improvements in tools supporting the B method.

% [Trabalhos futuros ] LLVM:Existe também a pretensão de especificar as
In the future, we plan to improve the support for the development of
software with the B method from functional specification to assembly
level, using the Z80 model presented in this work. For instance, the
mechanic compilation from B algorithmic constructs to assembly
platform is also envisioned. Finally, another important activity is to
develop a formal model of a platform used in LLVM
Compiler~\cite{DBLP:conf/cgo/LattnerA04}, which would enable us to
integrate different compilation techniques into our B-based framework.

% Conclusão: toque final
%Several iterations were needed to provide the good library definitions as well as
%to fine-tune the model of the microcontroller instructions by factoring common
%functionalities into auxiliary definitions.
